---
title: "Automated Cell Type Annotation: Comparing Azimuth, SingleR, and scType"
subtitle: "Source: `cell-type-annotation.qmd`"
author: "Steve Rozen (sr110@duke.edu, steverozen@pm.me)"
date: last-modified
date-format: "YYYY-MM-DD HH:mm"
execute:
  warning: false
format:
  html:
    page-layout: full
    include-in-header:
      text: |
        <style>
          .quarto-container, .page-columns, .main-container {
            max-width: 100% !important;
          }
        </style>
---

## Introduction

Manual cell type annotation relies on expert knowledge of canonical markers and
can be subjective. Automated methods offer reproducibility, speed, and the
ability to leverage large reference datasets. This document compares three
approaches on the PBMC3K dataset from the
[clustering tutorial](pbmc3k-tutorial.html):

- **Azimuth** — reference-mapping via Seurat's anchor-based transfer learning,
  using a pre-built PBMC reference atlas
- **SingleR** — correlation-based annotation against reference transcriptomes
  of purified cell types
- **scType** — marker-gene-based scoring using a curated database of cell type
  markers

## Setup

```{r}
#| label: setup
#| message: false
#| warning: false
# Installation (run once, not part of this document):
# install.packages("Azimuth", repos = "https://satijalab.r-universe.dev")
# BiocManager::install(c("SingleR", "celldex", "SingleCellExperiment"))
# install.packages("ggalluvial")
# scType: no install needed (sourced from GitHub at runtime)

library(Seurat)
library(Azimuth)
library(SingleR)
library(celldex)
library(SingleCellExperiment)
library(dplyr)
library(ggplot2)
library(patchwork)
library(ggalluvial)
library(HGNChelper)

base_size <- 14
theme_set(theme_bw(base_size = base_size))
```

## Load Data

Load the processed PBMC3K Seurat object from the clustering tutorial. This
object has clusters 0--8, UMAP embedding, and manual cell type labels in
`Idents()`.

```{r}
#| label: load-data
pbmc <- readRDS(here::here("output", "pbmc3k_final.rds"))
pbmc
```

Store the manual annotations in a dedicated metadata column so they are
preserved when `Idents()` is overwritten by downstream methods.

```{r}
#| label: store-manual-labels
pbmc$manual_annotation <- Idents(pbmc)
knitr::kable(as.data.frame(table(CellType = pbmc$manual_annotation)),
             row.names = FALSE)
```

## Method 1: Azimuth (Reference Mapping)

Azimuth maps query cells onto a pre-built reference atlas (Hao et al. 2021)
using anchor-based transfer learning. It provides hierarchical labels at two
resolutions (L1 broad, L2 fine) along with prediction scores indicating
annotation confidence.

Note: the first run downloads the reference (~1.5 GB).

```{r}
#| label: azimuth-run
#| message: false
#| warning: false
pbmc <- RunAzimuth(pbmc, reference = "pbmcref")
```

```{r}
#| label: azimuth-explore
head(pbmc@meta.data[, c("predicted.celltype.l1", "predicted.celltype.l2",
                         "predicted.celltype.l1.score",
                         "predicted.celltype.l2.score")])
```

### Prediction score distributions

Higher scores indicate greater annotation confidence.

```{r}
#| label: azimuth-scores
#| fig-width: 10
#| fig-height: 4
p1 <- ggplot(pbmc@meta.data, aes(x = predicted.celltype.l1.score)) +
  geom_histogram(bins = 50) +
  ggtitle("Azimuth L1 Prediction Scores")
p2 <- ggplot(pbmc@meta.data, aes(x = predicted.celltype.l2.score)) +
  geom_histogram(bins = 50) +
  ggtitle("Azimuth L2 Prediction Scores")
p1 + p2
```

### UMAP by Azimuth labels

```{r}
#| label: azimuth-umap
#| fig-width: 14
#| fig-height: 6
p1 <- DimPlot(pbmc, group.by = "predicted.celltype.l1",
              label = TRUE, repel = TRUE) +
  NoLegend() + ggtitle("Azimuth L1")
p2 <- DimPlot(pbmc, group.by = "predicted.celltype.l2",
              label = TRUE, repel = TRUE) +
  NoLegend() + ggtitle("Azimuth L2")
p1 + p2
```

## Method 2: SingleR (Reference Correlation)

SingleR correlates each cell's expression profile against reference datasets
of purified cell types. We use the Monaco Immune Data reference, which
provides both broad (`label.main`) and fine-grained (`label.fine`)
annotations for immune cells.

```{r}
#| label: singler-reference
#| message: false
monaco_ref <- MonacoImmuneData()
```

Convert the Seurat object to SingleCellExperiment format for SingleR input.

```{r}
#| label: singler-convert
#| warning: false
# SeuratObject/alabaster.base S4 class cache and reddim_names warnings are
# harmless artifacts of the Seurat-to-SCE conversion
sce <- as.SingleCellExperiment(pbmc)
```

```{r}
#| label: singler-run
#| message: false
#| warning: false
singler_main <- SingleR(
  test   = sce,
  ref    = monaco_ref,
  labels = monaco_ref$label.main
)

singler_fine <- SingleR(
  test   = sce,
  ref    = monaco_ref,
  labels = monaco_ref$label.fine
)
```

```{r}
#| label: singler-add-metadata
pbmc$singler_monaco_main <- singler_main$labels
pbmc$singler_monaco_fine <- singler_fine$labels
```

### Annotation confidence

SingleR prunes low-confidence assignments (setting them to `NA`).

```{r}
#| label: singler-diagnostics
data.frame(
  Level = c("Main", "Fine"),
  `Pruned cells` = c(sum(is.na(singler_main$pruned.labels)),
                      sum(is.na(singler_fine$pruned.labels))),
  check.names = FALSE
) |> knitr::kable(row.names = FALSE)
```

### UMAP by SingleR labels

```{r}
#| label: singler-umap
#| fig-width: 14
#| fig-height: 6
p1 <- DimPlot(pbmc, group.by = "singler_monaco_main",
              label = TRUE, repel = TRUE) +
  NoLegend() + ggtitle("SingleR (Monaco, main)")
p2 <- DimPlot(pbmc, group.by = "singler_monaco_fine",
              label = TRUE, repel = TRUE) +
  NoLegend() + ggtitle("SingleR (Monaco, fine)")
p1 + p2
```

## Method 3: scType (Marker-Gene Scoring)

scType scores cells using a curated marker gene database without needing a
reference expression dataset. Unlike Azimuth and SingleR (which annotate
per cell), scType assigns one label per cluster.

```{r}
#| label: sctype-source
#| message: false
source("https://raw.githubusercontent.com/IanevskiAleksandr/sc-type/master/R/gene_sets_prepare.R")
source("https://raw.githubusercontent.com/IanevskiAleksandr/sc-type/master/R/sctype_score_.R")
```

```{r}
#| label: sctype-db
#| warning: false
# checkGeneSymbols warns about non-standard gene symbol casing in the scType DB
db_url <- paste0(
  "https://raw.githubusercontent.com/IanevskiAleksandr/sc-type/",
  "master/ScTypeDB_full.xlsx"
)
tissue_type <- "Immune system"
gs_list <- gene_sets_prepare(db_url, tissue_type)
```

```{r}
#| label: sctype-score
DefaultAssay(pbmc) <- "RNA"

es_max <- sctype_score(
  scRNAseqData = pbmc[["RNA"]]$scale.data,
  gs  = gs_list$gs_positive,
  gs2 = gs_list$gs_negative
)

# Aggregate scores per cluster and pick the top-scoring cell type
cL_results <- do.call("rbind", lapply(
  levels(pbmc$seurat_clusters),
  function(cl) {
    es_cl <- sort(
      rowSums(es_max[, rownames(pbmc@meta.data[pbmc$seurat_clusters == cl, ])]),
      decreasing = TRUE
    )
    head(data.frame(
      cluster = cl,
      type    = names(es_cl),
      score   = es_cl,
      ncells  = sum(pbmc$seurat_clusters == cl)
    ), 3)
  }
))

sctype_labels <- cL_results |>
  group_by(cluster) |>
  dplyr::filter(score == max(score)) |>
  arrange(as.integer(as.character(cluster)))

knitr::kable(sctype_labels, row.names = FALSE, digits = 1)
```

```{r}
#| label: sctype-add-metadata
pbmc$sctype_annotation <- sctype_labels$type[
  match(pbmc$seurat_clusters, sctype_labels$cluster)
]
```

### UMAP by scType labels

```{r}
#| label: sctype-umap
#| fig-width: 10
#| fig-height: 7
DimPlot(pbmc, group.by = "sctype_annotation",
        label = TRUE, repel = TRUE) +
  NoLegend() + ggtitle("scType")
```

## Comparison of Methods

### Side-by-side UMAPs

```{r}
#| label: comparison-umap-grid
#| fig-width: 14
#| fig-height: 12
p_manual  <- DimPlot(pbmc, group.by = "manual_annotation",
                     label = TRUE, repel = TRUE) +
  NoLegend() + ggtitle("Manual Annotation")
p_azimuth <- DimPlot(pbmc, group.by = "predicted.celltype.l1",
                     label = TRUE, repel = TRUE) +
  NoLegend() + ggtitle("Azimuth L1")
p_singler <- DimPlot(pbmc, group.by = "singler_monaco_main",
                     label = TRUE, repel = TRUE) +
  NoLegend() + ggtitle("SingleR (Monaco)")
p_sctype  <- DimPlot(pbmc, group.by = "sctype_annotation",
                     label = TRUE, repel = TRUE) +
  NoLegend() + ggtitle("scType")

(p_manual + p_azimuth) / (p_singler + p_sctype)
```

### Cross-tabulations

#### Manual vs. Azimuth L1

```{r}
#| label: cross-tab-azimuth
knitr::kable(
  as.data.frame.matrix(table(
    Manual = pbmc$manual_annotation,
    Azimuth = pbmc$predicted.celltype.l1
  ))
)
```

#### Manual vs. SingleR (Monaco, main)

```{r}
#| label: cross-tab-singler
knitr::kable(
  as.data.frame.matrix(table(
    Manual = pbmc$manual_annotation,
    SingleR = pbmc$singler_monaco_main
  ))
)
```

#### Manual vs. scType

```{r}
#| label: cross-tab-sctype
knitr::kable(
  as.data.frame.matrix(table(
    Manual = pbmc$manual_annotation,
    scType = pbmc$sctype_annotation
  ))
)
```

### Proportional heatmap (Manual vs. Azimuth)

```{r}
#| label: agreement-heatmap
#| fig-width: 10
#| fig-height: 7
cross_df <- as.data.frame(table(
  Manual  = pbmc$manual_annotation,
  Azimuth = pbmc$predicted.celltype.l1
))

cross_df <- cross_df |>
  group_by(Manual) |>
  mutate(Proportion = Freq / sum(Freq))

ggplot(cross_df, aes(x = Azimuth, y = Manual, fill = Proportion)) +
  geom_tile() +
  geom_text(aes(label = Freq),
            size = 0.7 * base_size / ggplot2::.pt) +
  scale_fill_gradient(low = "white", high = "steelblue") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = rel(0.8))) +
  ggtitle("Manual vs. Azimuth L1 Agreement")
```

### Alluvial plot

This shows how cells flow between annotation schemes. Only flows with
`r 5` or more cells are shown for readability.

```{r}
#| label: alluvial-plot
#| fig-width: 12
#| fig-height: 8
#| warning: false
annotation_df <- data.frame(
  Manual  = as.character(pbmc$manual_annotation),
  Azimuth = as.character(pbmc$predicted.celltype.l1),
  SingleR = as.character(pbmc$singler_monaco_main)
)

alluvial_df <- annotation_df |>
  dplyr::count(Manual, Azimuth, SingleR) |>
  dplyr::filter(n >= 5)

ggplot(alluvial_df,
       aes(axis1 = Manual, axis2 = Azimuth, axis3 = SingleR, y = n)) +
  geom_alluvium(aes(fill = Manual), width = 1/12) +
  geom_stratum(width = 1/12, fill = "grey90", color = "grey50") +
  geom_text(stat = "stratum",
            aes(label = after_stat(stratum)),
            size = 0.55 * base_size / ggplot2::.pt) +
  scale_x_discrete(limits = c("Manual", "Azimuth", "SingleR")) +
  theme(legend.position = "none") +
  ggtitle("Cell Type Label Flow Across Methods")
```

## Discussion

All three methods largely agree on the major PBMC cell types (T cells,
monocytes, B cells, NK cells), which is expected for a well-characterized
tissue with abundant reference data.

Key differences between the methods:

- **Azimuth** provides hierarchical labels (L1/L2) and prediction scores,
  making it easy to assess confidence. It requires a matching reference atlas,
  which limits applicability to tissues with available references.

- **SingleR** annotates per-cell against bulk reference transcriptomes,
  offering broad reference coverage via `celldex`. It provides pruning of
  low-confidence assignments. Results are sensitive to the choice of reference
  dataset.

- **scType** uses marker gene databases and assigns one label per cluster
  (not per cell). It is transparent and interpretable but depends on the
  quality of the marker database and inherits the granularity of the
  upstream clustering.

For novel or poorly characterized tissues where reference atlases are
unavailable, SingleR with a broad reference or scType with custom marker
lists may be the only viable automated options.

## Save Annotated Object

```{r}
#| label: save-annotated
saveRDS(pbmc, file = here::here("output", "pbmc3k_annotated.rds"))
```

## References

- Hao Y et al. (2024) "Dictionary learning for integrative, multimodal and scalable single-cell analysis." *Nat Biotechnol* 42:293--304. <https://doi.org/10.1038/s41587-023-01767-y>
- Hao Y et al. (2021) "Integrated analysis of multimodal single-cell data." *Cell* 184:3573--3587. <https://doi.org/10.1016/j.cell.2021.04.048>
- Aran D et al. (2019) "Reference-based analysis of lung single-cell sequencing reveals a transitional profibrotic macrophage." *Nat Immunol* 20:163--172. <https://doi.org/10.1038/s41590-018-0276-y>
- Monaco G et al. (2019) "RNA-Seq Signatures Normalized by mRNA Abundance Allow Absolute Deconvolution of Human Immune Cell Types." *Cell Rep* 26:1627--1640. <https://doi.org/10.1016/j.celrep.2019.01.041>
- Ianevski A, Giri AK, Aittokallio T (2022) "Fully-automated and ultra-fast cell-type identification using specific marker combinations from single-cell transcriptomic data." *Nat Commun* 13:1246. <https://doi.org/10.1038/s41467-022-28803-w>

::: {.callout-note collapse="true"}
## Session Info

```{r}
#| label: session-info
#| echo: false
#| comment: ""
sessioninfo::session_info(info = "all")
```

:::
